---
title: "Mortgage Market EDA"
output: html_notebook
---

```{r include = FALSE}
# Clear environment of variables and functions
rm(list = ls(all = TRUE)) 

# Clear environmet of packages
if(is.null(sessionInfo()$otherPkgs) == FALSE)lapply(paste("package:", names(sessionInfo()$otherPkgs), sep=""), detach, character.only = TRUE, unload = TRUE)
```

#Load Packages 

```{r}
library(readr)
library(tidyverse)
library(janitor)
library(kableExtra)
library(gridExtra)
```

#Load and manipulate data 
```{r}
#Load Data 
mld <- read_csv("MLD Data File-1.csv")

```

```{r}
#Data Manipulation 
names(mld) = c("married", "guideline", "obligation", "black", "hispanic", "male", "approve", "loan") 

mld$male <- as.numeric(mld$male)
mld$married <- as.numeric(mld$married)

summary(mld)
```

#Sample Selection Criteria 

* There are 12 NAs in Male and 3 NA in Married. Given our data set of almost 2,000 observation, we decided to remove these observations.
* There are values of 666 in "Guideline", which is an impossible value. We decided to remove such observations as well. 

```{r include = FALSE}
#Data Manipulation 
#Remove NAs and impossible values 
mld<-mld %>% 
  filter(guideline != 666) %>% 
  filter(!is.na(married)) %>% 
  filter(!is.na(male))

#Rename Race 
mld$race <- ifelse((mld$black == 0) & (mld$hispanic == 0), "Non-Hispanic White", 
                   ifelse((mld$black == 1) & (mld$hispanic == 0), "Non-Hispanic Black", "Hispanic"))

#Rename Gender
mld$gender <-ifelse(mld$male == 1, "Male", "Female")
mld$marital <-ifelse(mld$married== 1, "Married", "Unmarried")

mld <- mld %>% select(-black, -hispanic, -male, -married)

#Change variables levels' name:
mld$approve <- ifelse(mld$approve == 1, "Yes", "No")
mld$guideline <- ifelse(mld$guideline == 1, "Yes", "No")
```

#Descriptive statistics

##All observations analysis 
```{r}
#Categorical Variables 
bar_graph <- function(x, xtitle) { 
  ggplot(mld, aes(x)) + geom_bar() +
    theme_bw() + xlab(xtitle)+
    theme(panel.border = element_blank(), 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())
}
grid.arrange(
bar_graph(mld$marital, "Marital Status"),
bar_graph(mld$approve, "Approval Rate"),
bar_graph(mld$race, "Race") + coord_flip(), 
bar_graph(mld$gender, "Gender"))

#Continuous Variables
par(mfrow = c(1,2))
hist(mld$loan)
hist(mld$obligation)

```


##Descriptive statistics by race/ethnicity
```{r}
kable(
  mld %>% 
    tabyl(race, approve) %>% 
    adorn_totals(where = "col") %>%
    adorn_percentages(denominator="row")%>%
      adorn_pct_formatting(digits = 0 ),
      digits = 2,
      format = "html",
      align = c("l","c","c"), 
  caption = "Approval Rate by Race/Ethnicity"
    )%>%
  kable_styling(full_width = F)

mld%>% group_by(race) %>% summarise(mean_loan = mean(loan), 
                                    mean_obl = mean(obligation))
  
```

